Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
a number of fixes

\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

@@

Tracing JITs are an idea initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM
\cite{andreas_gal_trace-based_2009} and has been tried for Adobe's Tamarin
ActionScript VM \cite{chang_tracing_2009}.

@@

The basic approach of a tracing JIT is to only generate machine code for the hot
code paths of commonly executed loops and to interpret the rest of the program.
The code for those common loops however should be highly optimized, including
aggressive inlining.

Typically, programs executed by a tracing VMs goes through various phases:
\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
\toon{next sentence is strange} To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers. A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

Since the tracing JIT cannot know which parts of the language interpreter are
the program counter, the author of the language interpreter needs to mark the
relevant variables of the language interpreter with the help of a \emph{hint}.
The tracing interpreter will then effectively add the values of these variables
to the position key. This means that the loop will only be considered to be
closed if these variables that are making up program counter at the language
interpreter level are the same a second time.  Loops found in this way are, by
definition, user loops.

@@

\toon{very difficult to read (actually so is the whole paragraph; rephrase)}
The simple insight how to improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are ``green''), that means that the tracer checks that they
are some fixed value at the beginning of the loop (they may well change over the
course of the loop, though). In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It is possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
that when no hints were present in the interpreter at all. \toon{strange
sentence}

@@

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython. \sout{by a bit. The tracer needs to
trace through quite a bit of dispatching machinery of the Python interpreter to
achieve this, XXX.}
\anto{
To achieve this, the tracer traces through the whole Python dispatching
machinery, automatically inlining only the relevant fast paths.
}

>>>>

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

@@

Tracing JITs are an idea initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006, andreas_gal_incremental_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM
\cite{andreas_gal_trace-based_2009} and has been tried for Adobe's Tamarin
ActionScript VM \cite{chang_tracing_2009}.

@@

The basic approach of a tracing JIT is to only generate machine code for the hot
code paths of commonly executed loops and to interpret the rest of the program.
The code for those common loops however is highly optimized, including
aggressive inlining.

Typically, programs executed by a tracing VM go through various phases:
\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}

The tracer interprets these functions in a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. The trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator.  The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers.\footnote{The
chain of \texttt{if}, \texttt{elif}, ... instructions that check the various
opcodes is transformed into a \texttt{switch} statement by one of PyPy's
optimizations. Python does not have a \texttt{switch} statement}
A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

Since the tracing JIT cannot know which parts of the language interpreter are
the program counter, the author of the language interpreter needs to mark the
relevant variables of the language interpreter with the help of a \emph{hint}.
The tracing interpreter will then effectively add the values of these variables
to the position key. This means that the loop will only be considered to be
closed if these variables that are making up the program counter at the language
interpreter level are the same a second time.  Loops found in this way are, by
definition, user loops.

@@

The simple insight how to improve the situation is that most of the operations
in the trace are actually concerned with manipulating the bytecode string and
the program counter. Those are stored in variables that are ``green'' (e.g. they
are part of the position key).  This means that the tracer checks that those
variables have some fixed value at the beginning of the loop (they may well
change over the course of the loop, though). In the example of Figure
\ref{fig:trace-no-green-folding} the check would be that at the beginning of the
trace the \texttt{pc} variable is \texttt{4} and the \texttt{bytecode} variable
is the bytecode string corresponding to the square function. Therefore it is
possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It is possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process.

@@

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython. To achieve this, the tracer
traces through the whole Python dispatching machinery, automatically inlining
the relevant fast paths.
